Modeling the competing effects of the immune system and EMT on epithelial cancers

### 2019-12-02 Paper/CleanPaper.tex
finished this round of edits!

The majority of deaths from cancer are due to metastasis of the disease \cite{dillekas}. It is thus of critical importance to understand better the progression from in situ to invasive disease. Underlying this progression are genetic and epigenetic events, including mutations in pathways critical to the success of the cancer cell (driver mutations) \cite{ryan2016hallmarks}. These pathways include cell proliferation, apoptosis, and immunogenicity.

@@

@@

@@

The other facet of mesenchymal cells is their slower proliferation rate, and this is tied to the observation that mesenchymal cells have been speculate to exhibit a more ``stemness'' phenotype.

@@

Mathematical oncology, that is, mathematical models of cancer incidence, progression, and treatment, has become a well-developed field; many models have offered insight into the cellular interactions underlying cancer and its interplay with the immune system, including classic \cite{anderson98_continuous, sherrattjonathana.92_oncogenes, pillis05_validated} and more recent works \cite{kim18_cell, gallaher14_bridging, gallaher18_spatial, an15_agentbased, serre16_mathematical, louzoun14_mathematical, briones-orta13_arkadia, lavi13_role, greene15_modeling, greene16_mathematical, cho17_modeling-1,  benzekry17_mathematical, owen11_mathematical, west18_multidrug}. These studies have increased our understanding of how tumors grow in the presence of various immune components, and how treatment regimes can be designed to maximize the efficacy of cytotoxicity while minimizing risks to the patient. However, to our knowledge no models have addressed how the effects of EMT alter interactions between the immune system and cancer, and the subsequent implications for treatment. 
\par
Here we develop a model with the goal of studying interactions between the tumor, the immune system and EMT.
We seek to describe a set of crucial molecular and cellular interactions in epithelial tumor cells, including effects due to DNA damage and mutation, to investigate the probability that in situ tumors will progress and, if so, when.
A recent model of cancer-immune interactions \cite{guo17_multiscale} described the effects of the TME on the risk of cancer, and we build on the core cell cycle component of this model, adding significant new interactions to the immune component of the model (which was previously modeled by a single interaction), as well as adding the effects of EMT.
In doing so, we shift the focus of the previous model from cancer initiation to cancer progression.
We do this to reflect the fact that cancer progression hinges on escape from the immune system and the fact that EMT has a more well-defined role during progression and metastasis.
We seek to understand whether this more complex immune module will change our understanding of inflammatory effects on the tumor, and how the epithelial-mesenchymal axis influences these.
\par 
In the next section we develop the model and give explanation of the intuition behind each component.
We then go on to analyze its behavior.
Global sensitivity analysis identifies parameters that are crucial for progression.
We then study these in more depth, focusing on the competing effects of EMT and of the immune system on progression: we find that EMT intricately regulates progression and that under certain regimes a careful balance of EMT- and immune-driven processes can significantly prolong cancer-free survival.
To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects of inflammation and EMT predicted by the model for patients with pancreatic cancer.

@@

\subsection{EMT}\label{EMT}
Each tissue cell has an EMT score between 0 and 1.
When the score is above a fixed threshold value, the cell is labeled as mesenchymal.
Otherwise, it is below the threshold and labeled as epithelial.
For the purpose of parameter values of the model, a cell in an epithelial state is considered in the base state, and one in a mesenchymal state will have a subset of its parameters updated.
In particular, mesenchymal cells benefit from increased immune evasion but suffer from decreased proliferation.
Note, that because the immune system can only clear malignant cells, the increased immune evasion only benefits malignant mesenchymal cells.
\par
In modeling EMT this way, we are assuming that the same factor, TGF-$\beta$, drives EMT both at initiation and through progression of cancer.
In addition, as per our model assumptions (see Table \ref{table:model_assumptions}), we choose to ignore other factors that are known to induce EMT such as hypoxia as well as the typical spatial organization of epithelial and mesenchymal cells as this pertains to the invasive front.
\par
For the cells that have undergone EMT and are currently labeled in the model as mesenchymal, they experience two modifications to their governing parameters.
The first is a reduction in proliferation, which we call mesenchymal growth arrest (MGA).
This causes to mesenchymal cells to be less likely to undergo proliferation and more likely to rest during a cell cycle.
The second is a decrease to their likelihood of being cleared by the immune system, which we call mesenchymal immune evasion (MIE). 
This decreases the likelihood of mesenchymal cells being cleared by both NKs and CTLs during a cell cycle.
Though we have not found clinical data on which to base parameter values for either of these, by their very nature they are proportions on the interval $[0,1]$.
Thus, it is possible to sample from the entire parameter and in fact we do present results from a large subset of this parameter space and our sensitivity analysis fully explores this space.

@@

\subsubsection{Tumor cell fate}
During each cell cycle, every cell is randomly assigned one of four fates: proliferation, apoptosis, immune clearance, and rest in $G_0$.
Depending on both cell intrinsic, e.g. EMT and mutational state, and cell extrinsic, e.g. immune infiltrate and total number of tumor cells, each of the four options is given a probability of occurring for each cell.
For proliferation, we assume the total number of tumor cells exerts a negative feedback on the proliferation probability for each cell.
Also, cells that have a driver mutation in the proliferation pathway have an increased probability of proliferating.
Meanwhile, mesenchymal cells have a decreased probability of proliferating and an increased probability of resting.
For apoptosis, the probability is constant except for those cells with a driver mutation in the apoptosis pathway which have a decreased probability of apoptosis.
For immune clearance, we assume that only cells with newly acquired driver mutations are targeted by the immune system, and their probability of being cleared depends on the immune infiltrate.
This probability can be decreased by a cell having a driver mutation in the immune evasion pathway or by a cell being mesenchymal.
Further details and equations are given in the supplement.

\subsubsection{Completing the Cell Cycle}
Once all the cell fates are carried out, the rest of the components of the model update to account for these changes.
First, the immune populations are updated in two steps.
In step one, Immune cells are exhausted based on the number of malignant cells that were cleared by their type.
So, each clearance by an NK cell results in the NK population decreasing by one, and similarly for CTLs.
In step two, all three immune populations are updated deterministically based on a simple ODE model of their dynamics.
NKs have a constant source and death rate while CTLs and Tregs have a source rate that is dependent on the number of malignant cells cleared that cycle and a constant death rate.
In addition, the concentration of TGF-$\beta$ mediates recruitment of Tregs with higher concentrations of TGF-$\beta$ resulting in a greater influx of Tregs.
\par
Also at this time, new driver mutations have a chance to occur.
The cells that proliferated are the only ones that have a probability of acquiring a new driver mutation.
This probability is cell-specific and is increased if no mutation occurs and reset to 0 if a mutation does occur.
\par
Finally, the concentration of TGF-$\beta$ is updated as well as the EMT score for each cell is updated.
Malignant cells and Tregs are the two sources of TGF-$\beta$ each producing a constant amount.
Tregs are assumed to produce much more TGF-$\beta$ than malignant cells.
The resulting quantity of TGF-$\beta$ is divided up randomly among the tumor cells.
This quantity is combined with the EMT score of each cell and the resulting quantity determines which cells undergo EMT and which undergo MET.
The equations governing these processes can be found in the supplement.

@@

We begin by investigating general features of the model to establish baseline conditions and to assess the impact of various model components on the key measured outcomes: the probability of progression, and the Time to Cancer. 
Within the cell cycle, cell fate is determined via a set of rules that are influenced by EMT and immune interactions (Fig. \ref{fig:ModelIntro}A). For example, if a cell undergoes EMT, the probability that it will proliferate is reduced; if a cell gains a driver mutation in the apoptosis pathway, the probability that it undergoes apoptosis is greatly reduced.
The different means by which the immune system acts on tumor cells are also shown (\ref{fig:ModelIntro}A Inset). NK cells and CTLs attempt to clear malignant tumor cells, and deactivate upon successfully carrying out their cytotoxic function.
Tregs inhibit cytotoxic activity.
Tregs also release TGF-$\beta$ which promotes the recruitment of Tregs, as well as increasing the rate of EMT.
\par
The inflammation cycling scheme for a typical {\it in silico} patient consists of alternating high and low regimes (Fig. \ref{fig:ModelIntro}B); the inflammation schemes modeled will be discussed in detail in the next section.
For this patient, after the warmup period, cell mutations are observed at a rate low enough that they are cleared by cytotoxic cells before being able to establish a tumor.
After about 700 cell cycles, the malignant cell population begins to grow steadily.
At this point, large numbers of cells from the adaptive immune system (CTLs and Tregs) are being quickly recruited into the TME, and a peak in TGF-$\beta$ expression is observed.
After 841 cell cycles, the proportion of malignant cells reaches 50\%: the threshold defining progression.
Thus this patient has a Time to Cancer of 841 cell cycles, or 631 days.
For this patient, we kept let the simulation go beyond this timepoint, and we se that the number of malignant cells grows rapidly and soon makes up 100\% of the cell population.
A peak in number of mesenchymal cells is observed shortly after this threshold is reached, and following this most cells transition back to epithelial.
\par 
Considering the immune system dynamics, we see that the NK population is approximately constant, while the adaptive populations (CTL and Treg cells) grow quickly and dwarf NK cells in number following the accumulation of driver mutations.
The adaptive immune populations also appear to exhibit oscillatory behavior, however note that this is not due to intrinsic dynamics but rather due to the inflammation scheme that the patient is undergoing: alternating between 30 cell cycles of high inflammation and 60 cell cycles of low inflammation. Within each of these periods, adaptive immune populations increase or decrease rapidly in accordance with the inflammation state.
\par 
The {\em in silico} patient described here is given for the purpose of illustrating features of the model. Given the multiple sources of noise in the model, in order to quantify patient dynamics and cancer-free survival rates, below we will simulate large cohorts of patients. In Fig. \ref{fig:ModelIntro}C we simulate the survival curve for a cohort of $500$ patients: we see that all the patients survive for approximately 100 cell cycles (75 days). Subsequently, in the approximate range of $T= [100,300]$, the cancer onset rate is roughly constant, and after $600$ cell cycles, no patients remain cancer free.

\subsection{Identifying regulatory parameters via Morris global sensitivity analysis}\label{SensAnalysis}
Exploring the parameter spaces of models in systems biology is -- in general -- a hard problem. Performing Bayesian parameter inference to inform parameter values is advisable wherever possible \cite{kirk13_model}. Here, a lack of detailed molecular measurements  (i.e. data for tumor growth dynamics abound, but simultaneous data on the immune dynamics are lacking) preclude inference of the full model. In addition, while inference schemes for agent-based models are developing \cite{gallaher17_hybrid, warne19_simulation}, simulation times remain a hurdle \cite{lambert18_bayesian}. Parameters for aspects of this model that were studied previously can be constrained \cite{guo17_multiscale}, however even for these, the new additions to the model could push it into new behavioral regimes. Thus to adequately sample the parameter space of the model and identify those areas of parameter space that exert the most control, we use sensitivity analysis.   
\par
To assess the sensitivity of the model parameters and thus identify those that are most important in determining cancer-free survival times, we performed Morris one-step-at-a-time (OAT) sensitivity analysis (see Methods). 
The results of Morris OAT on the 31 model parameters are shown in Figure \ref{fig:MOAT}. We see that a subset of parameters demonstrate much higher levels of sensitivity than others.
The two most influential according to this analysis are the death rate and the recruitment rate of Treg cells, this is most likely due to the dual roles Treg cells play in both suppressing the cytotoxic effects of other immune cells and secreting TGF-$\beta$, which drives EMT.
This ties Treg cells to all three components of the model.
Since we seek to separate the effects of different model components, we do not choose the parameters influencing Treg cells for detailed analysis below.
\par 
Many parameters in the model change depending on the inflammation state.
For the sake of a naming convention, any parameter which includes ``low'' in the name represents the parameter value when inflammation is low.
For when the inflammation is high, these ``low'' parameters get scaled by parameters which include ``up'' in the name.
The ILD and IHD parameters determine the duration of low and high inflammation, respectively.
A cell with altered pathways will be affected by the $\Delta_A$, $\Delta_P$, and $\Delta_{\text{IE}}$ parameters, depending on whether the apoptosis, proliferation, or immune evasion pathways are altered.
The mesenchymal cells are influenced by the MGA and MIE parameters.
Finally, $\sigma$, $k_\text{EMT}$, $\tau_\text{max}$, and $K_3$ all control how cells transition between the epithelial and mesenchymal states.

@@

When a cell transitions from an epithelial to a mesenchymal state, two phenotypic cell characteristics change: mesenchymal immune evasion (MIE) and mesenchymal growth arrest (MGA).
Both parameters are defined proportionally, and lie in $[0,1]$, where higher values indicate more mesenchymal-like properties.
The parameter $\Delta_\text{MIE}$ is the proportional reduction in the probability that an invasive mesenchymal cell will be subject to immune clearance in a given cycle.
The parameter $\Delta_\text{MGA}$ is the proportional reduction in the probability that a mesenchymal cell will proliferate in a given cell cycle, thus is equivalent (strictly in terms of its impact on the cell cycle) to an increased proportion of time spent resting in the $G_0$ phase.
The cytokine TGF-$\beta$ is also involved in the EMT process (by increasing the probability of EMT). All three of these parameters were found to have large effects on the model by the Morris OAT analysis presented in Section \ref{SensAnalysis}.
\par 
As MIE increases, the invasion-free survival decreases (Fig. \ref{fig:FirstSurvivalCurves}A) under all sets of parameters studied: as this subpopulation of invasive cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
These results are summarized by Fig. \ref{fig:FirstSurvivalCurves}A.
\par
As MGA increases, invasion-free survival increases: lower proliferation rates for mesenchymal cells slow down cancer progression (Fig \ref{fig:FirstSurvivalCurves}B).
This is not intuitive, since decreased proliferation rates affects all mesenchymal cells, not just the invasive cells.
\par
Third, TGF-$\beta$ can be varied in two ways: the production by mesenchymal cells and the production by Treg cells.
In Fig \ref{fig:FirstSurvivalCurves}C, the results of varying Treg TGF-$\beta$ production are shown, indicating that an increased Treg TGF-$\beta$ production leads to a shorter invasion-free survival time.
The two main ways in which TGF-$\beta$ influences the system is in recruitment of Treg cells and in pushing tumor cells to a mesenchymal phenotype.
Treg cells are modeled as tumor-protective and thus increasing their number will naturally decrease the invasion-free survival.
Mesenchymal cells are more likely to evade the immune system, so pushing the system towards an overall more mesenchymal state will lead to poorer outcomes, decreasing invasion-free survival times.

@@

We explored the effects of varying the inflammation state of the patient on invasion-free survival, to investigate competing interactions within the tumor microenvironment and their effect on EMT. Patient cohorts were simulated under different inflammation regimes: permanently low inflammation; permanently high inflammation; or variable inflammation. For patients drawn from cohorts in a permanently high inflammatory state, the relationship between mesenchymal parameters and invasion-free survival is monotonic, i.e. increasing either MIE (Fig. \ref{fig:VaryINFL_and_MesPars}A-B) or MGA (Fig. \ref{fig:VaryINFL_and_MesPars}C-D) decreases the invasion-free survival time.
However, under regimes with either temporary or permanent periods of low inflammation, different relationships emerge: a local maximum for the invasion-free survival time is found with respect to MGA ($\Delta_\text{MGA}= 0.3$). This is seen both for intermittent or low inflammation (Fig. \ref{fig:VaryINFL_and_MesPars}D). 
\par
These striking differences in the mean invasion-free survival -- that can be extended by up to one year by optimization of the mesenchymal growth arrest -- have clear therapeutic implications. 
Our model predicts that a patient suffering intermittent inflammatory attacks at the tumor site will benefit from controlling the proliferation of the mesenchymal cells.
\par
In contrast, when MIE is varied for different inflammation cycling schemes, regardless of the cycling scheme, increasing the MIE will decreases the invasion-free survival (worsen cancer progression and prognosis). In the case where the inflammatory state is permanently high, however, MIE has little effect on the invasion-free survival time. Thus under any inflammation regime with periods of low inflammation, reductions of any size in the rate of mesenchymal immune evasion lead to improved patient outcomes.

@@

We found that this model recapitulated invasion-free survival dynamics, and via global parameter sensitivity analysis, we identified parameters exerting key control over model behavior. Focusing on these led us to identify that: increasing mesenchymal immune evasion, decreasing mesenchymal growth arrest, or increasing Treg TGF-$\beta$ production all lead to shorter invasion-free survival times. However, varying the level of inflammation led to paradoxical effects: under regimes with periods of low inflammation, an optimal level of mesenchymal growth arrest can improve outcomes and maximize the invasion-free survival.

>>>>

The majority of deaths from cancer are due to metastasis of the disease \cite{dillekaas201990}. It is thus of critical importance to understand better the progression from in situ to invasive disease. Underlying this progression are genetic and epigenetic events, including mutations in pathways critical to the success of the cancer cell (driver mutations) \cite{ryan2016hallmarks}. These pathways include cell proliferation, apoptosis, and immunogenicity.

@@

%%%%%%% I think we should reorder these next two paragraphs

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The other facet of mesenchymal cells is their slower proliferation rate, and this is tied to the observation that mesenchymal cells have been speculated to exhibit a more ``stemness'' phenotype.

@@

Mathematical oncology, that is, mathematical models of cancer incidence, progression, and treatment, has become a well-developed field; many models have offered insight into the cellular interactions underlying cancer and its interplay with the immune system, including classic \cite{anderson98_continuous, sherrattjonathana.92_oncogenes, pillis05_validated} and more recent works \cite{kim18_cell, gallaher14_bridging, gallaher18_spatial, an15_agentbased, serre16_mathematical, louzoun14_mathematical, briones-orta13_arkadia, lavi13_role, greene15_modeling, greene16_mathematical, cho17_modeling-1,  benzekry17_mathematical, owen11_mathematical, west18_multidrug}. These studies have increased our understanding of how tumors grow in the presence of various immune components, and how treatment regimes can be designed to maximize the efficacy of cytotoxicity while minimizing risks to the patient. However, to our knowledge no models have addressed how the effects of EMT alter interactions between the immune system and cancer, and the subsequent implications for treatment. 
Here we develop a model with the goal of studying interactions between the tumor, the immune system and EMT.
We seek to describe a set of crucial molecular and cellular interactions in epithelial tumor cells, including effects due to DNA damage and mutation, to investigate the probability that in situ tumors will progress and, if so, when.
A recent model of cancer-immune interactions \cite{guo17_multiscale} described the effects of the TME on the risk of cancer, and we build on the core cell cycle component of this model, adding significant new interactions to the immune component of the model (which was previously modeled by a single interaction), as well as adding the effects of EMT.
In doing so, we shift the focus of the previous model from cancer initiation to cancer progression.
We do this to reflect the fact that cancer progression hinges on escape from the immune system and the fact that EMT has a more well-defined role during progression and metastasis.
We seek to understand whether this more complex immune module will change our understanding of inflammatory effects on the tumor, and how the epithelial-mesenchymal axis influences these.
In the next section we develop the model and give explanation of the intuition behind each component.
We then go on to analyze its behavior.
Global sensitivity analysis identifies parameters that are crucial for progression.
We then study these in more depth, focusing on the competing effects of EMT and of the immune system on progression: we find that EMT intricately regulates progression and that under certain regimes a careful balance of EMT- and immune-driven processes can significantly prolong cancer-free survival.
To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects of inflammation and EMT predicted by the model for patients with pancreatic cancer.

@@

\subsection{EMT}\label{EMT}
Each tissue cell has an EMT score between 0 and 1.
When the score is above a fixed threshold value, the cell is labeled as mesenchymal.
Otherwise, it is below the threshold and labeled as epithelial.
For the purpose of parameter values of the model, a cell in an epithelial state is considered in the base state, and one in a mesenchymal state will have a subset of its parameters updated.
In particular, mesenchymal cells benefit from increased immune evasion but suffer from decreased proliferation.
Note, that because the immune system can only clear malignant cells, the increased immune evasion only benefits malignant mesenchymal cells.
\par
In modeling EMT this way, we are assuming that the same factor, TGF-$\beta$, drives EMT both at initiation and through progression of cancer.
% Do we still want the table of assumptions with the quick guide box?
%In addition, as per our model assumptions (see Table \ref{table:model_assumptions}), we choose to ignore other factors that are known to induce EMT such as hypoxia as well as the typical spatial organization of epithelial and mesenchymal cells as this pertains to the invasive front.
In addition, we choose to ignore other factors that are known to induce EMT such as hypoxia as well as the typical spatial organization of epithelial and mesenchymal cells as this pertains to the invasive front.
\par
For the cells that have undergone EMT and are currently labeled in the model as mesenchymal, they experience two modifications to their governing parameters.
The first is a reduction in proliferation, which we call mesenchymal growth arrest (MGA).
This causes to mesenchymal cells to be less likely to undergo proliferation and more likely to rest during a cell cycle.
The second is a decrease to their likelihood of being cleared by the immune system, which we call mesenchymal immune evasion (MIE). 
This decreases the likelihood of mesenchymal cells being cleared by both NKs and CTLs during a cell cycle.
Though we have not found clinical data on which to base parameter values for either of these, by their very nature they are proportions on the interval $[0,1]$.
Thus, it is possible to sample from the entire parameter and in fact we do present results from a large subset of this parameter space and our sensitivity analysis fully explores this space.

@@

\subsubsection{Tumor cell fate}
During each cell cycle, every cell is randomly assigned one of four fates: proliferation, apoptosis, immune clearance, and rest in $G_0$.
Depending on both cell intrinsic, e.g. EMT and mutational state, and cell extrinsic, e.g. immune infiltrate and total number of tumor cells, each of the four options is given a probability of occurring for each cell.
For proliferation, we assume the total number of tumor cells exerts a negative feedback on the proliferation probability for each cell.
Also, cells that have a driver mutation in the proliferation pathway have an increased probability of proliferating.
Meanwhile, mesenchymal cells have a decreased probability of proliferating and an increased probability of resting.
For apoptosis, the probability is constant except for those cells with a driver mutation in the apoptosis pathway which have a decreased probability of apoptosis.
%For immune clearance, we assume that only cells with newly acquired driver mutations are targeted by the immune system, and their probability of being cleared depends on the immune infiltrate.
For immune clearance, the malignant tumor cells are assumed more immunogenic and thus have a higher probability of being cleared by the immune system.
This probability can be decreased by a cell having a driver mutation in the immune evasion pathway or by a cell being mesenchymal.
Further details and equations are given in the supplement.

\subsubsection{Completing the Cell Cycle}
Once all the cell fates are carried out, the rest of the components of the model update to account for these changes.
First, the immune populations are updated in two steps.
In step one, immune cells are exhausted based on the number of malignant cells that were cleared by their type.
So, each clearance by an NK cell results in the NK population decreasing by one, and similarly for CTLs.
In step two, all three immune populations are updated deterministically based on a simple ODE model of their dynamics.
NKs have a constant source and death rate while CTLs and Tregs have a source rate that is dependent on the number of malignant cells cleared that cycle and a constant death rate.
In addition, the concentration of TGF-$\beta$ mediates recruitment of Tregs with higher concentrations of TGF-$\beta$ resulting in a greater influx of Tregs.
\par
Also at this time, new driver mutations have a chance to occur.
The cells that proliferated are the only ones that have a probability of acquiring a new driver mutation.
This probability is cell-specific and is increased if no mutation occurs and reset to 0 if a mutation does occur.
\par
Finally, the concentration of TGF-$\beta$ is updated as well as the EMT score for each cell is updated.
Malignant cells and Tregs are the two sources of TGF-$\beta$ each producing a constant amount.
Tregs are assumed to produce much more TGF-$\beta$ than malignant cells.
The resulting quantity of TGF-$\beta$ is divided up randomly among the tumor cells.
This quantity is combined with the EMT score of each cell and the resulting quantity determines which cells undergo EMT and which undergo MET.
The equations governing these processes can be found in the supplement.

@@

We begin by investigating general features of the model to establish baseline conditions and to assess the impact of various model components on the key measured outcomes: the probability of progression, and the Time to Cancer. 
Within the cell cycle, cell fate is determined via a set of rules that are influenced by EMT and immune interactions (Fig. \ref{fig:ModelIntro}A). For example, if a cell undergoes EMT, the probability that it will proliferate is reduced; if a cell gains a driver mutation in the apoptosis pathway, the probability that it undergoes apoptosis is greatly reduced.
The different means by which the immune system acts on tumor cells are also shown (\ref{fig:ModelIntro}A Inset). NK cells and CTLs attempt to clear malignant tumor cells, and deactivate upon successfully carrying out their cytotoxic function.
Tregs inhibit cytotoxic activity.
Tregs also release TGF-$\beta$ which promotes the recruitment of Tregs, as well as increasing the rate of EMT.
\par
The inflammation cycling scheme for a typical {\it in silico} patient consists of alternating high and low regimes (Fig. \ref{fig:ModelIntro}B); the inflammation schemes modeled will be discussed in detail below.
For this patient, after the warmup period, cell mutations are observed at a rate low enough that they are cleared by cytotoxic cells before being able to establish a tumor.
After about 700 cell cycles, the malignant cell population begins to grow steadily.
At this point, large numbers of cells from the adaptive immune system (CTLs and Tregs) are being quickly recruited into the TME, and a peak in TGF-$\beta$ expression is observed.
After 841 cell cycles, the proportion of malignant cells reaches 50\%: the threshold defining progression.
Thus this patient has a Time to Cancer of 841 cell cycles, or 631 days.
For this patient, we let the simulation go beyond this timepoint, and we see that the number of malignant cells grows rapidly and soon makes up 100\% of the cell population.
A peak in number of mesenchymal cells is observed shortly after this threshold is reached, and following this most cells transition back to epithelial.
\par 
Considering the immune system dynamics, we see that the NK population is approximately constant, while the adaptive populations (CTL and Treg cells) grow quickly and dwarf NK cells in number following the accumulation of driver mutations.
The adaptive immune populations also appear to exhibit oscillatory behavior, however note that this is not due to intrinsic dynamics but rather due to the inflammation scheme that the patient is undergoing: alternating between 30 cell cycles of high inflammation and 30 cell cycles of low inflammation. Within each of these periods, adaptive immune populations increase or decrease rapidly in accordance with the inflammation state.
\par 
The {\em in silico} patient described here is given for the purpose of illustrating features of the model. Given the multiple sources of noise in the model, in order to quantify patient dynamics and cancer-free survival rates, below we will simulate large cohorts of patients. In Fig. \ref{fig:ModelIntro}C we simulate the survival curve for a cohort of $500$ patients: 
%we see that all the patients survive for approximately 100 cell cycles (75 days). Subsequently, in the approximate range of $T= [100,300]$, the cancer onset rate is roughly constant, and after $600$ cell cycles, no patients remain cancer free.
we see that after the mutagenic event that a large number of tumors quickly progress while others takes some time before progressing. By about 1200 cell cycles, all tumors have progressed.

\subsection{Identifying regulatory parameters via Morris global sensitivity analysis}\label{SensAnalysis}
Exploring the parameter spaces of models in systems biology is -- in general -- a hard problem. Performing Bayesian parameter inference to inform parameter values is advisable wherever possible \cite{kirk13_model}. Here, a lack of detailed molecular measurements  (i.e. data for tumor growth dynamics abound, but simultaneous data on the immune dynamics are lacking) preclude inference of the full model. In addition, while inference schemes for agent-based models are developing \cite{gallaher17_hybrid, warne19_simulation}, simulation times remain a hurdle \cite{lambert18_bayesian}. Parameters for aspects of this model that were studied previously can be constrained \cite{guo17_multiscale}, however even for these, the new additions to the model could push it into new behavioral regimes. Thus to adequately sample the parameter space of the model and identify those areas of parameter space that exert the most control, we use sensitivity analysis.   
\par
To assess the sensitivity of the model parameters and thus identify those that are most important in determining progression-free survival times, we performed Morris one-step-at-a-time (OAT) sensitivity analysis (see Methods). 
The results of Morris OAT on the 31 model parameters are shown in Figure \ref{fig:MOAT}. We see that a subset of parameters demonstrate much higher levels of sensitivity than others.
%The two most influential according to this analysis are the death rate and the recruitment rate of Treg cells, this is most likely due to the dual roles Treg cells play in both suppressing the cytotoxic effects of other immune cells and secreting TGF-$\beta$, which drives EMT.
The two most influential according to this analysis are the recruitment rates of Tregs and CTLs under low inflammation.
%This ties Treg cells to all three components of the model.
%Since we seek to separate the effects of different model components, we do not choose the parameters influencing Treg cells for detailed analysis below.
We also note that the parameters related to EMT have a moderate effect on Time to Cancer.
\par 
Many parameters in the model change depending on the inflammation state.
For the sake of a naming convention, any parameter which includes ``low'' in the name represents the parameter value when inflammation is low.
For when the inflammation is high, these ``low'' parameters get scaled by parameters which include ``up'' in the name.
The ILD and IHD parameters determine the duration of low and high inflammation, respectively.
A cell with altered pathways will be affected by the $\Delta_A$, $\Delta_P$, and $\Delta_{\text{IE}}$ parameters, depending on whether the apoptosis, proliferation, or immune evasion pathways are altered.
The mesenchymal cells are influenced by the MGA and MIE parameters.
Finally, $\sigma$, $k_\text{EMT}$, $\tau_\text{max}$, and $K_3$ all control how cells transition between the epithelial and mesenchymal states.

@@

When a cell transitions from an epithelial to a mesenchymal state, two phenotypic cell characteristics change: mesenchymal immune evasion (MIE) and mesenchymal growth arrest (MGA).
Both parameters are defined proportionally and hence lie in $[0,1]$, where higher values indicate more mesenchymal-like properties.
The parameter $\Delta_\text{MIE}$ is the proportional reduction in the probability that an invasive mesenchymal cell will be subject to immune clearance in a given cycle.
The parameter $\Delta_\text{MGA}$ is the proportional reduction in the probability that a mesenchymal cell will proliferate in a given cell cycle, thus is equivalent (strictly in terms of its impact on the cell cycle) to an increased proportion of time spent resting in the $G_0$ phase.
The cytokine TGF-$\beta$ is also involved in the EMT process (by increasing the probability of EMT). All three of these parameters were found to have large effects on the model by the Morris OAT analysis presented in Section \ref{SensAnalysis}.
\par 
As MIE increases, the invasion-free survival decreases (Fig. \ref{fig:FirstSurvivalCurves}A) under all sets of parameters studied: as this subpopulation of invasive cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
These results are summarized by Fig. \ref{fig:FirstSurvivalCurves}D.
\par
The relationship between MGA and invasion-free survival is non-monotonic.
For smaller values of MGA, an increase in MGA results in increased invasion-free survival (Fig \ref{fig:FirstSurvivalCurves}E).
This result is explored in Section \ref{KeyEMT}.
%As MGA increases, invasion-free survival increases: lower proliferation rates for mesenchymal cells slow down cancer progression (Fig \ref{fig:FirstSurvivalCurves}B).
%This is not intuitive, since decreased proliferation rates affects all mesenchymal cells, not just the invasive cells.
\par
Third, TGF-$\beta$ can be varied in two ways: the production by malignant cells and the production by Treg cells.
In Fig \ref{fig:FirstSurvivalCurves}F, the results of varying Treg TGF-$\beta$ production are shown, indicating that an increased Treg TGF-$\beta$ does not have much effect on invasion-free survival time.
The effect we do see is that lower Treg TGF-$\beta$ results in more rapid invasion or if it survives to a certain time point, a longer time to invasion.
Here, we choose to focus on the results of the effect of MIE and MGA as the more direct link to the mesenchymal phenotype.

@@

We explored the effects of varying the inflammation state of the patient on invasion-free survival, to investigate competing interactions within the tumor microenvironment and their effect on EMT. Patient cohorts were simulated under different inflammation regimes: permanently low inflammation; permanently high inflammation; or variable inflammation. 
Compared to the other inflammation regimes, a permanently high inflammation state results in tumors that are relatively resistant to changes in mesenchymal parameters (Fig. \ref{fig:VaryINFL_and_MesPars}).
However, when the inflammation either has temporary or permanent periods of low inflammation, different relationships emerge: invasion-free survival time becomes negatively correlated with MIE, and a local maximum for the invasion-free survival time is found with respect to MGA ($\Delta_\text{MGA}= 0.2$).

%For patients drawn from cohorts in a permanently high inflammatory state, the relationship between mesenchymal parameters and invasion-free survival is monotonic, i.e. increasing either MIE (Fig. \ref{fig:VaryINFL_and_MesPars}A-B) or MGA (Fig. \ref{fig:VaryINFL_and_MesPars}C-D) decreases the invasion-free survival time.
%However, under regimes with either temporary or permanent periods of low inflammation, different relationships emerge: a local maximum for the invasion-free survival time is found with respect to MGA ($\Delta_\text{MGA}= 0.3$). This is seen both for intermittent or low inflammation (Fig. \ref{fig:VaryINFL_and_MesPars}D). 
\par
These striking differences in the mean invasion-free survival -- that can be extended by up to double compared to other MGA values-- have clear therapeutic implications. 
Our model predicts that a patient suffering intermittent inflammatory attacks at the tumor site will benefit from controlling the proliferation of the mesenchymal cells.
\par
In contrast, when MIE is varied for different inflammation cycling schemes, regardless of the cycling scheme, increasing the MIE will decreases the invasion-free survival (worsen cancer progression and prognosis) or have minimal effect. Thus under any inflammation regime with periods of low inflammation, reductions of any size in the rate of mesenchymal immune evasion lead to improved patient outcomes.

@@

We found that this model recapitulated invasion-free survival dynamics, and via global parameter sensitivity analysis, we identified parameters exerting key control over model behavior. Focusing on these led us to identify that: increasing mesenchymal immune evasion and increasing Treg TGF-$\beta$ production both lead to shorter invasion-free survival times. However, varying the level of inflammation led to paradoxical effects with regards to mesenchymal growth arrest: under regimes with periods of low inflammation, an optimal level of mesenchymal growth arrest can improve outcomes and maximize the invasion-free survival.
